Theories of deliberated choice

### 2023-06-02 simple.tex
Introduced then revised depth using iterated defense

While numerous articles have discussed the concept of deliberation in the last decades, it remains unsettled, to the best of my knowledge, how to observe a deliberated judgment in a systematic and reproducible fashion. There is a need to define this concept sufficiently precisely that, first, it is clear what counts as a deliberated judgment (i.e., when “sufficient” \citep{meinard_justification_2020} or “correct” deliberation has happened); and second, it is clear which part of the claim is empirical and which part stems from an unfalsifiable conception of what “deliberated” means. This permits to study deliberated judgments empirically, meaning here that disagreements about someone’s deliberated judgment in a given context can be solved, at least in principle, by an empirical experiment, provided the very conception of what constitutes a deliberated judgment is shared. This article is interested in defining the concept of deliberated judgement of a subject and show how one can study it in an emprical manner. The “classical” counterpart of deliberated judgments,, that does not involve confrontation with arguments, will henceforth be called the natural judgment. As we deal with choice acts, we will also name the concept “deliberated choice”, as opposed to “natural choice”.

While the concept of deliberation in the deliberative democracy literature usually refers to multiple individuals debating, this article considers deliberation as related to a single individual at a time: an individual deliberates (in the sense of: carefully weights) the stance she considers most appropriate while being confronted to arguments. The deliberated judgment of an individual is thus defined with no requirement that the arguments be uttered by peers in a debate; they can be given by an automata, for example. In this sense, deliberated judgments come close to what \citet{rawls_theory_1999} calls reflective equilibrium (referring to an idea of \citet{goodman_fact_1983}), a concept that has much inspired the present proposal.

@@

Studying judgments resulting from exposure to arguments includes a difficulty that is not present with natural choice experiments: exposing an individual to arguments may change the individual’s stance, an effect that cannot be erased in order to submit the individual to an unrelated sequence of arguments. The observer thus cannot in general assume that multiple unrelated sequences of arguments can be tested on a given individual. This difficulty will be worked around by considering general theories of deliberated judgments, that apply to sets of individuals. This is methodologically similar to defining theories of how objects break. One can test only one way of breaking a given (unrepairable) object, so such theories must apply to sets of objects if one wants to be able to test multiple ways of falsifying the theories.

@@

\subsection{Exposure protocol}
An exposure protocol is a tuple $(I, \allargs, {\allleadsto})$, where $I$ is a set of individuals, $\allargs$ is a set of arguments, and ${\allleadsto}$ is the behavior of the individuals as a function of the arguments they have been exposed to.

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path. Its length is denoted by $\card{\hist} \in \N$ and its elements, $\hist(k)$ or $\hist_k$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist(k) = \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

The function ${\allleadsto}: I × \allhist → P$ associates to each individual and path an element of $P$, ${\allleadsto}(i, \alpha)$ representing the choice of $i$ following exposure to the arguments of $\alpha$, in order. That is, ${\allleadsto}(i, \alpha) = \phi$ denotes that $i$ chooses the first act after exposure to $\alpha$, ${\allleadsto}(i, \alpha) = ¬\phi$ denotes $i$ choosing the second act and ${\allleadsto}(i, \alpha) = 0$ is used to represent indifference of $i$.
The notation $\alpha \ileadsto p$ will be used to mean ${\allleadsto}(i, \alpha) = p$, considering ${\ileadsto} \subseteq \allhist × P$ as a binary relation.
As usual with binary relations, $\hist \nileadsto p$ means ${\allleadsto}(i, \alpha) ≠ p$ or equivalently ${\allleadsto}(i, \alpha) \in P \setminus \set{p}$.

@@

Given $i \in I, p \in P, \ar \in \allargs$, define $\ar \ileadstost p$ iff $\forall \hist \in \allhist \suchthat \ar \in \histend: \hist \ileadsto p$.
When $\ar \ileadstost p$, $\ar$ is said to be a decisive argument for $\prop$ from $i$’s point of view given the exposure protocol.

@@

This latter possibility may look unpleasant, but it is a fact that facing some difficult problems, individuals sometimes are undecided, and it is an a priori empirical possibility that the individual remains undecided (or even, becomes \emph{more} undecided) when considering multiple arguments.

\subsection{An example and a discussion about observability}
The following toy example is inspired by an argument from \citet[p.~59]{railton_facts_2003}. Here is his argument.
\begin{quote}
An experiment in cognitive psychology reveals that when subjects observe two individuals taking a test, one of whom answers a high percentage of the early questions correctly but then falters, while the other does
poorly initially but then answers a high percentage of the later questions
correctly, the first is usually viewed as more able, even though in the end
each answers the same number of questions correctly in all. If the subjects had been asked which test-taker they would prefer as a math tutor,
they likely would have answered “The first,” and perhaps they would also
have been willing to pay a premium to secure his services. But if they
were to become convinced of the influence of sheer order on their evaluations of relative ability, they presumably would no longer want their
initial preference for the first test-taker to influence their choices – or
willingness to pay – in this way.
\end{quote}
\begin{example}[Math tutor]
	\label{ex:tutor}
	Let $I$ be a set of individuals desiring to evaluate two candidates $\set{m_1, m_2}$ for math tutoring.
	Let ${\leadsto}(i, a_1)$ represent $i$ being shown a test of candidate $m_1$ wherein $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly.
	Let ${\leadsto}(i, a_2)$ represent $i$ being shown a test of candidate $m_2$ wherein $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly.
%	Let ${\leadsto}(i, a_e)$ represent $i$ being given an explanation about the psychological phenomenon described in the quote above.
	Let ${\leadsto}(i, a_e)$ represent $i$ being given the following argument. “When tested, $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly; whereas $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly. It follows than $m_1$ answers on average less well than $m_2$. Observing the candidates take the tests may yield a contrary impression; this is a well-known psychological phenomenon depending on the order of the questions that the candidates answer correctly.”
	Let $\prop_1$ represent the choice of $m_1$ and $\prop_2$ that of $m_2$.

	If the prediction described in the quote is correct when applied to this modified experiment, then, given $i \in I$, $i$’s deliberated choice is $\prop_2$ because $a_e$ is a decisive argument for $\prop_2$, even though $i$ might have rather chosen $\prop_1$ if given only the arguments $a_1$ and $a_2$ without $a_e$.
\end{example}

That the deliberated choice of any individual $i$ is well-defined may be doubtful, as it involves counterfactuals.
To see this, let $\alpha = (a_1, a_2, a_e)$ designate the argumentative path where the arguments are shown in the order they are described in the example, and let $\alpha' = (a_e, a_2, a_1)$ designate the reverse order.
The deliberated choice concept is defined so as to not depend on the (arbitrary) order of the arguments shown to $i$: for $\prop \in P_i$, it must be that directing to $i$ the path $\alpha$ or the path $\alpha'$ both lead $i$ to choose $\prop$.
However, once we have shown $\alpha$ to $i$, we cannot show to her $\alpha'$ (unless assuming $i$ forgets arguments after a while and waiting for sufficiently long, which would create practical and theoretical difficulties).

This difficulty is resolved using falsifiable general theories, applicable to $I$ and not solely to a given $i$, so that $\alpha$ can be tested on $i$ and $\alpha'$ on some $i' ≠ i$. Indeed, what is needed is a way to make a claim such as the one concluding \cref{ex:tutor}, according to which $a_e$ is a decisive argument, empirically testable. This is achieved in the next section thanks to theories of deliberated choice.
% Such theories contain an observable claim and a conclusion claim, so as to satisfy two properties. First, such a theory must be falsifiable in principle, meaning that it must be possible, using our observational basis (propositions of the sort $\alpha \ileadsto \phi$), to invalidate the observable claim of the theory. As a counter-example, using \cref{ex:tutor}, a theory that claims that $\forall i \in I: \alpha_{123} \ileadsto \prop_1 \lor \alpha_{321} \ileadsto \prop_1$ is not falsifiable in principle, as, whatever the real $\ileadsto$ relation, with a given $i$, one can never show that $\alpha_{123} \nileadsto \prop_1 \land \alpha_{123} \nileadsto \prop_1$. Second, such a theory must be so that if its observable claim holds, then some given alternative is in the deliberated judgment of the individuals in $I$.

@@

\begin{definition}[Static theory]
	\label{def:static}
	A static theory is a pair $(p, \ar)$ for some $p \in P$ and some $\ar \in \allargs$.
\end{definition}
A static theory $(p, \ar)$ is said to be truthful iff $\forall i \in I: \ar \ileadstost p$, and to be correct iff $\forall i \in I: p \in P_i$.
We can then say that such theories are valid, meaning that their truthfulness implies their correctness. Let us make this (very simple) result stand out as a theorem, because a parallel will be usefully made with more complicated theories.
%Note that this result immediately follows from \cref{ax:decisive}.
\begin{theorem}[Static theories are valid]
	Any truthful static theory is correct.
\end{theorem}
\begin{proof}
	This follows immediately from \cref{def:decisive}.
\end{proof}
Importantly, truthfulness can indeed be checked using our observational basis, that is, if $\exists i \in I \suchthat \ar \nileadstost p$, then it is possible to exhibit this fact experimentally.

Such theories have the merit of simplicity, but there is a need to design more adaptive theories: in many interesting cases, no single argument will be considered decisive by all individuals in $I$. What we need is a way to argue with individuals by reacting to what they consider good arguments. The following section illustrates this need.

@@

In many decision problems, individuals will not be attracted by the same set of arguments and counter-arguments, even though they may finally opt for the same alternative. Imagine that some individual $i_1$ considers an argument $a_1$ for $\nprop$ relevant and some individual $i_2$ considers an argument $a_2$ for $\nprop$ relevant, but $i_1$ does not consider $a_2$ relevant and $i_2$ does not consider $a_1$ relevant (for example, because they already have some counter-arguments in mind to those arguments). Imagine further that there is some counter-argument $a'_1$ that can be given to $i_1$ to rebut $a_1$, and a different counter-argument $a'_2$ that can be given to $i_2$ to rebut $a_2$, so that they both have $\prop$ in their deliberated judgment when considering all these arguments.
In such situation, static theories as defined in \cref{sec:static} are ill suited: these theories need to present the same argument to all individuals and would thus need to define a bigger argument $a$ containing both $a'_1$ and $a'_2$. Such an argument could be too long to be convincing to $i_1$ or $i_2$, as it needlessly spends efforts countering an argument that the individual considers irrelevant to start with. This  will become an important problem if there are not only two but many a priori possible arguments $a_k$ for $\nprop$, a small subset of which is considered relevant by any given individual, each with their own counter-argument $a'_k$ for $\prop$.

%For a partial illustration about an ambitious setting, consider a decision to be taken about the relevance of some concrete action to help cope with climate change, and let $I$ be a set of citizens interested in the climate issue. The complete set of all arguments that will be considered relevant by at least one individual may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ against the proposed concrete action, it is possible to give counter-arguments $A'$ that will convince the individual to opt for the proposal.
One of the settings discussed by \citet{thaler_nudge_2009} permits to illustrate this point more concretely.
Let $I$ be a set of workers interested in establishing a savings plan. 
The authors claim that workers tend to not save enough because of psychological biases. 
It is therefore tempting to conjecture that the deliberated choice of those workers is to save more, in the sense defined here, than what their natural choice (the one revealed in absence of arguments) leads to.
A difficulty related to this decision problem is that the complete set of all arguments that will be considered relevant by at least one individual for such a complex issue may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ for saving not much (i.e., spending more in the short term), it is possible to give counter-arguments $A'$ that will convince the individual to opt for saving more.
But no single argument could be decisive for each $i \in I$: it could be that only counter-arguments built specifically according to what a given individual considers relevant could convince him, given the complexity of the issue.
This illustrates that the deliberated choice of each individuals in $I$ may be to save more even though no static theory (\cref{def:static}) may achieve showing it, as it could be that no such theory is truthful.

@@

The following example illustrates this same point in a constrained setting which permits a complete definition of $\allargs$ and $\leadsto$.%, which permits defining the setting completely at the price of a reduced realism.
\begin{example}[A decision problem about two lotteries]
	\label{ex:lichtenstein}
	In a famous article about preference reversals, \citet{lichtenstein_reversals_2006} discuss an example involving two lotteries. The lottery $P$ lets the subject win four dollars with 99\% chance and lose ten cents with 1\% chance; in summary, $L_P = (.99, +\$4; .01, −\$0.1)$. The lottery $L_D$ is, using the same notations: $(.33, +\$16; .67, −\$2)$. The letter $P$ stands for probability: one salient aspect of $L_P$ is the very high chance of winning. The letter $D$ stands for dollar: one salient aspect of $L_D$ is the high gain it gives (compared to $P$) in dollar amount if the subject is lucky.

@@

Define $I$ as a set of individuals facing a choice between $L_P$ and $L_D$ in a laboratory experiment.
Define $\ileadsto$ as simply exposing $i$ to $\alpha$ and observing the resulting choice by asking $i$ to indicate which lottery she prefers playing. 
%Given an argumentative path $\hist$, $\hist \ileadsto (P ≥ D)$ iff, in a situation of $i$ having been exposed, in order, to the arguments in $\hist$, letting $i$ choose between $P$ and $D$ or declare indifference leads $i$ to choose $P$, or declare indifference. Here is a possible text presented to $i$ to ask for a choice: “Considering these arguments, would you please choose either to play $P$, or play $D$, or declare to be indifferent (in which case one will be picked randomly for you)”. 
%Similarly, define $\hist \ileadsto (P > D)$ iff showing the arguments in $\hist$ to $i$ and letting $i$ choose between $P$ and $D$ or declare indifference leads $i$ to choose $P$ and not $D$. 
%Define similarly $\hist \ileadsto \prop$ for $\prop \in \set{D ≥ P, D > P}$.

@@

\subsection{Argumentative theories of deliberated choice}
\label{sec:dynamic}
An argumentative theory of deliberated choice $\gamma$ is defined as a tuple $(\pga, \arga, \dg)$, where $\pga \in P$ is the proposition supported by the theory, $\arga \in \allargs$ is the initial argument in favor of $\pga$ and $\dg$ is a defense relation.

The defense relation $\dg \subseteq (\allargs × \allhist) × \allargs$ defines the argumentation strategy of the theory. Informally speaking, its role is, together with $\arga$, to produce the arguments that the theory claims will convince the individuals in $I$ to choose $\pga$. Assuming that the theory has played an argument $\ar_0$ and a path $\hist$ challenges its intended conclusion in favor of $\pga$ (as judged by the individual), $\dg(\ar_0, \hist)$ is the argument that the theory plays as a counter-argument to the path $\hist$.

Formally speaking, a defense relation $\dg \subseteq (\allargs × \allhist) × \allargs$ is a right-unique relation satisfying bounded depth (to be defined shortly). Right-unique means that given any $\ar \in \allargs, \hist \in \allhist$, $\dg(\ar, \hist)$ is empty or is a singleton. %, and must satisfy $\ar \notin \histend ⇒ \dg(\ar, \hist) = \emptyset$. 
As is usual for right-unique relations, we write $\dg(\ar_0, \hist) = \ar_1$ instead of $\dg(\ar_0, \hist) = \set{\ar_1}$.
Say that $\ar_1$ defends $\ar_0$ iff $\exists \hist \suchthat \dg(\ar_0, \hist) = \ar_1$.
Define $\dg(\ar) = \dg(\ar, \allhist) = \set{\dg(\ar, \hist) \suchthat \hist \in \allhist} \subseteq \allargs$ as all the defenders of $\ar$.

This mechanism permits to define defenders of defenders of an argument, $\dg(\dg(\ar))$, defenders of defenders of defenders of an argument, … This yields the following notion of depth.
Given any set $S$ and function $f: S → S$, let $f^0$ represent the identity function, and given $k \in \N$, let $f^{k + 1} = f \circ f^k$.
The bounded depth condition on the defense relation imposes that
$\exists k \in \N^* \suchthat (\dg)^k(\arga) = \emptyset$.
The smallest such $k$ is called the depth of the theory.

%The condition that when $\dg(\ar_0, \hist)$, $\ar_0 \in \histend$ is imposed for simplicity: as will appear, an argument $\ar_0$ needs defense only when it is challenged by some path $\hist \suchthat \ar_0 \in \histend$, thus, letting $\dg$ define counter-arguments to paths not satisfying this condition would be of no use.

>>>>

While numerous articles have discussed the concept of deliberation in the last decades, it remains unsettled, to the best of my knowledge, how to observe a deliberated judgment in a systematic and reproducible fashion. There is a need to define this concept sufficiently precisely that, first, it is clear what counts as a deliberated judgment (i.e., when “sufficient” \citep{meinard_justification_2020} or “correct” deliberation has happened); and second, it is clear which part of the claim is empirical and which part stems from an unfalsifiable conception of what “deliberated” means. This permits to study deliberated judgments empirically, meaning here that disagreements about someone’s deliberated judgment in a given context can be solved, at least in principle, by an empirical experiment, provided the very conception of what constitutes a deliberated judgment is shared. This article is interested in defining the concept of deliberated judgement of a subject and show how one can study it in an empirical manner. The “classical” counterpart of deliberated judgments,, that does not involve confrontation with arguments, will henceforth be called the natural judgment. As we deal with choice acts, we will also name the concept “deliberated choice”, as opposed to “natural choice”.

While the concept of deliberation in the deliberative democracy literature usually refers to multiple individuals debating, this article considers deliberation as related to a single individual: an individual deliberates (in the sense of: carefully weights) the stance she considers most appropriate while being confronted to arguments. The deliberated judgment of an individual is thus defined with no requirement that the arguments be uttered by peers in a debate; they can be given by an automaton, for example. In this sense, deliberated judgments come close to what \citet{rawls_theory_1999} calls reflective equilibrium (referring to an idea of \citet{goodman_fact_1983}), a concept that has much inspired the present proposal.

@@

Studying judgments resulting from exposure to arguments includes a difficulty that is not present with natural choice experiments: exposing an individual to arguments may change the individual’s stance, an effect that cannot be erased in order to submit the individual to an unrelated sequence of arguments. The observer thus cannot in general assume that multiple unrelated sequences of arguments can be tested on a given individual. This difficulty will be worked around by considering general theories of deliberated judgments, that apply to sets of individuals. This is methodologically similar to defining theories of how objects break. One can break a given (unrepairable) object only once, so such theories must apply to sets of objects if one wants to be able to test multiple ways of falsifying the theories.

@@

\subsection{Exposure protocol}
An exposure protocol is a tuple $(I, \allargs, {\allleadsto})$, where $I$ is a set of individuals, $\allargs$ is a set of arguments, and ${\allleadsto}$, “leads to”, is the behavior of the individuals as a function of the arguments they have been exposed to.

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N^*: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path. Its length is denoted by $\card{\hist} \in \N$ and its elements, $\hist(k)$ or $\hist_k$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist(k) = \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

The function ${\allleadsto}: I × \allhist → P$ associates to each individual and path an element of $P$, ${\allleadsto}(i, \alpha)$ representing the choice of $i$ following exposure to the arguments of $\alpha$, in order. That is, ${\allleadsto}(i, \alpha) = \phi$ denotes that $i$ chooses the first act after exposure to $\alpha$, ${\allleadsto}(i, \alpha) = ¬\phi$ denotes $i$ choosing the second act and ${\allleadsto}(i, \alpha) = 0$ is used to represent indifference of $i$.
The notation $\alpha \ileadsto p$, pronounced “$\alpha$ leads $i$ to $p$”, will be used to mean ${\allleadsto}(i, \alpha) = p$, considering ${\ileadsto} \subseteq \allhist × P$ as a binary relation.
As usual with binary relations, $\hist \nileadsto p$ means ${\allleadsto}(i, \alpha) ≠ p$ or equivalently ${\allleadsto}(i, \alpha) \in P \setminus \set{p}$.

@@

Given $i \in I, p \in P, \ar \in \allargs$, define $\ar \ileadstost p$, pronounced “$a$ leads definitely $i$ to $p$”, iff $\forall \hist \in \allhist \suchthat \ar \in \histend: \hist \ileadsto p$.
When $\ar \ileadstost p$, $\ar$ is said to be a decisive argument for $\prop$ from $i$’s point of view given the exposure protocol.

@@

This latter possibility may look unpleasant, but it is a fact that facing some difficult problems, individuals sometimes are undecided, and it is an a priori empirical possibility that the individual remains undecided (or even, \emph{becomes} undecided) when considering multiple arguments. 
Being undecided is not the same state as being decidedly indifferent about two issues, and those states can be distinguished empirically, as we will see.

\subsection{An example and a discussion about observability}
The following toy example is inspired by an argument from \citet[p.~59]{railton_facts_2003}. Here is his argument.
\begin{quote}
An experiment in cognitive psychology reveals that when subjects observe two individuals taking a test, one of whom answers a high percentage of the early questions correctly but then falters, while the other does
poorly initially but then answers a high percentage of the later questions
correctly, the first is usually viewed as more able, even though in the end
each answers the same number of questions correctly in all. If the subjects had been asked which test-taker they would prefer as a math tutor,
they likely would have answered “The first,” and perhaps they would also
have been willing to pay a premium to secure his services. But if they
were to become convinced of the influence of sheer order on their evaluations of relative ability, they presumably would no longer want their
initial preference for the first test-taker to influence their choices – or
willingness to pay – in this way.
\end{quote}
\begin{example}[Math tutor]
	\label{ex:tutor}
	Let $I$ be a set of individuals desiring to evaluate two candidates $\set{m_1, m_2}$ for math tutoring.
	Consider one individual $i \in I$.
	Let ${\leadsto}(i, a_1)$ represent $i$ being shown a test of candidate $m_1$ wherein $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly.
	Let ${\leadsto}(i, a_2)$ represent $i$ being shown a test of candidate $m_2$ wherein $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly.
%	Let ${\leadsto}(i, a_e)$ represent $i$ being given an explanation about the psychological phenomenon described in the quote above.
	Let ${\leadsto}(i, a_e)$ represent $i$ being given the following argument. “When tested, $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly; whereas $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly. It follows than $m_1$ answers on average less well than $m_2$. Observing the candidates take the tests may yield a contrary impression; this is a well-known psychological phenomenon depending on the order of the questions that the candidates answer correctly.”
	Let $\prop_1$ represent the choice of $m_1$ and $\prop_2$ that of $m_2$.

	If the prediction described in the quote applies to $i$, the relation $\ileadsto$ is as follows, $\forall k \in \N^*$ (recall that for $\args \subseteq \allargs$, $\args^k$ represents the set of sequences of length $k$ using arguments from $\args$): $\forall \hist \in \set{a_e}^k, \hist \ileadsto \prop_2$; $\forall \hist \in \set{a_1, a_2}^k, \hist \ileadsto \prop_1$; $\forall \hist \in \set{a_1, a_e}^k \suchthat a_e \in \histend, \hist \ileadsto \prop_2$; $\forall \hist \in \set{a_2, a_e}^k \suchthat a_e \in \histend, \hist \ileadsto \prop_2$; $\forall \hist \in \allargs^k \suchthat a_e \in \histend, \hist \ileadsto \prop_2$. The values of ${\leadsto}(i, \hist)$ for $\hist = \emptyset$ and for $\hist \in \allargs^k \suchthat a_e \notin \histend$ are undetermined by the prediction, but whatever they are, $a_e$ is a decisive argument for $\prop_2$, hence, $i$’s deliberated choice is $\prop_2$.
\end{example}

That the deliberated choice of any individual $i$ is well-defined may be doubtful, as it involves counterfactuals.
To see this, let $\hist = (a_1, a_2, a_e)$ designate the argumentative path where the arguments are shown in the order they are described in \cref{ex:tutor}, and let $\hist' = (a_e, a_2, a_1)$ designate the reverse order.
The deliberated choice concept is defined so as to not depend on the order of the arguments shown to $i$: for having $\prop \in P_i$, it must be that directing to $i$ the path $\hist$ or the path $\hist'$ both lead $i$ to choose $\prop$.
However, once we have shown $\hist$ to $i$, we cannot show to her $\hist'$ (unless assuming $i$ forgets arguments after a while and waiting for sufficiently long, which would create practical and theoretical difficulties).

This difficulty is resolved using falsifiable general theories, applicable to $I$ and not solely to a given $i$, so that $\hist$ can be tested on $i$ and $\hist'$ on some $i' ≠ i$. Indeed, we need a way to empirically test claims such as the one concluding \cref{ex:tutor}, according to which $a_e$ is a decisive argument. This is achieved in the next section thanks to theories of deliberated choice.
% Such theories contain an observable claim and a conclusion claim, so as to satisfy two properties. First, such a theory must be falsifiable in principle, meaning that it must be possible, using our observational basis (propositions of the sort $\alpha \ileadsto \phi$), to invalidate the observable claim of the theory. As a counter-example, using \cref{ex:tutor}, a theory that claims that $\forall i \in I: \alpha_{123} \ileadsto \prop_1 \lor \alpha_{321} \ileadsto \prop_1$ is not falsifiable in principle, as, whatever the real $\ileadsto$ relation, with a given $i$, one can never show that $\alpha_{123} \nileadsto \prop_1 \land \alpha_{123} \nileadsto \prop_1$. Second, such a theory must be so that if its observable claim holds, then some given alternative is in the deliberated judgment of the individuals in $I$.

@@

\begin{definition}[Static theory]
	\label{def:static}
	A static theory is a pair $(p, \ar)$ for some $p \in P$ and some $\ar \in \allargs$.
\end{definition}
A static theory $(p, \ar)$ is said to be truthful iff $\forall i \in I: \ar \ileadstost p$, and to be correct iff $\forall i \in I: p \in P_i$.
We can then deduce that such theories are valid, meaning that their truthfulness implies their correctness. Let us make this (very simple) result stand out as a theorem, because a parallel will be usefully made with more complicated theories.
%Note that this result immediately follows from \cref{ax:decisive}.
\begin{theorem}[Static theories are valid]
	Any truthful static theory is correct.
\end{theorem}
\begin{proof}
	This follows immediately from \cref{def:decisive}.
\end{proof}
Importantly, truthfulness can indeed be checked using our observational basis, that is, if $\exists i \in I \suchthat \ar \nileadstost p$, then it is possible to exhibit this fact experimentally.

Such theories have the merit of simplicity, but there is a need to design more adaptive theories: in many interesting cases, no single argument will be considered decisive by all individuals in $I$. We then need theories that argue with individuals by reacting to what they consider good arguments. The following section illustrates this need.

@@

In many decision problems, individuals will not be attracted by the same set of arguments and counter-arguments, even though they may finally settle to the same preference. Imagine that some individual $i_1$ considers an argument $a_1$ for $\nprop$ relevant and some individual $i_2$ considers an argument $a_2$ for $\nprop$ relevant, but $i_1$ does not consider $a_2$ relevant and $i_2$ does not consider $a_1$ relevant (for example, because they already have some counter-arguments in mind to those arguments). Imagine further that there is some counter-argument $a'_1$ that can be given to $i_1$ to rebut $a_1$, and a different counter-argument $a'_2$ that can be given to $i_2$ to rebut $a_2$, so that they both have $\prop$ in their deliberated judgment when considering all these arguments.
In such situation, static theories as defined in \cref{sec:static} are ill suited: these theories need to present the same argument to all individuals and would thus need to define a bigger argument $a$ containing both $a'_1$ and $a'_2$. Such an argument could be too long to be convincing to $i_1$ or $i_2$, as it needlessly spends efforts countering an argument that the individual considers irrelevant to start with. This  will become an important problem if there are not only two but many a priori possible arguments $a_k$ for $\nprop$, a small subset of which is considered relevant by any given individual, each with their own counter-argument $a'_k$ for $\prop$.
This section is devoted to describing two examples to illustrate that point. Bear in mind that these examples may look somehow naïve on the aspect of the hypothesis they make about what convinces individuals. Real-world complex decision situations will require complex theories that do not fit the purpose of simple illustrative examples. Their goal is to argue that non-static theories are required, but not to fully describe such theories.

%For a partial illustration about an ambitious setting, consider a decision to be taken about the relevance of some concrete action to help cope with climate change, and let $I$ be a set of citizens interested in the climate issue. The complete set of all arguments that will be considered relevant by at least one individual may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ against the proposed concrete action, it is possible to give counter-arguments $A'$ that will convince the individual to opt for the proposal.
One of the settings discussed by \citet{thaler_nudge_2009} permits such an illustration. %to illustrate this point more concretely.
Let $I$ be a set of workers interested in establishing a savings plan. 
The authors claim that workers tend to not save enough because of psychological biases. 
It is therefore tempting to conjecture that the deliberated choice of those workers is to save more, in the sense defined here, than what their natural choice (the one revealed in absence of arguments) leads to.
A difficulty related to this decision problem is that the complete set of all arguments that will be considered relevant by at least one individual for such a complex issue may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ for saving not much (i.e., spending more in the short term), it is possible to give counter-arguments $A'$ that will convince the individual to opt for saving more.
But no single argument could be decisive for each $i \in I$: it could be that only counter-arguments built specifically according to what a given individual considers relevant could convince her, given the complexity of the issue.
This illustrates that the deliberated choice of each individuals in $I$ may be to save more even though no static theory (\cref{def:static}) may achieve showing it, as it could be that no such theory is truthful.

@@

The following example illustrates this same point in a constrained setting which permits a complete definition of $\allargs$ and $\leadsto$.%, which permits defining the setting completely at the price of a reduced realism.
\begin{example}[A decision problem about two lotteries]
	\label{ex:lichtenstein}
	In a famous article about preference reversals, \citet{lichtenstein_reversals_2006} discuss an example involving two lotteries. The lottery $L_P$ lets the subject win four dollars with 99\% chance and lose ten cents with 1\% chance; in summary, $L_P = (.99, +\$4; .01, −\$0.1)$. The lottery $L_D$ is, using the same notations: $(.33, +\$16; .67, −\$2)$. The letter $P$ stands for probability: one salient aspect of $L_P$ is the very high chance of winning. The letter $D$ stands for dollar: one salient aspect of $L_D$ is the high gain it gives (compared to $P$) in dollar amount if the subject is lucky.

@@

Define $I$ as a set of individuals facing a choice between $L_P$ and $L_D$ in a laboratory experiment.
Define $\ileadsto$ as exposing $i$ to $\alpha$ and observing the resulting choice by asking $i$ to indicate which lottery she prefers playing, or declare being indifferent. 
%Given an argumentative path $\hist$, $\hist \ileadsto (P ≥ D)$ iff, in a situation of $i$ having been exposed, in order, to the arguments in $\hist$, letting $i$ choose between $P$ and $D$ or declare indifference leads $i$ to choose $P$, or declare indifference. Here is a possible text presented to $i$ to ask for a choice: “Considering these arguments, would you please choose either to play $P$, or play $D$, or declare to be indifferent (in which case one will be picked randomly for you)”. 
%Similarly, define $\hist \ileadsto (P > D)$ iff showing the arguments in $\hist$ to $i$ and letting $i$ choose between $P$ and $D$ or declare indifference leads $i$ to choose $P$ and not $D$. 
%Define similarly $\hist \ileadsto \prop$ for $\prop \in \set{D ≥ P, D > P}$.

@@

\subsection{Argumentative theories of deliberated choice}
\label{sec:dynamic}
An argumentative theory of deliberated choice $\gamma$ is defined as a tuple $(\pga, \dg)$, where $\pga \in P$ is the proposition supported by the theory and $\dg$ is a defense relation.

The defense relation $\dg \subseteq \allhist × \allargs$ defines the argumentation strategy of the theory. Informally speaking, it defines the arguments that the theory claims will convince the individuals in $I$ to choose $\pga$. Assuming that some path $\hist$ challenges the theory’s intended conclusion in favor of $\pga$ (as judged by the individual), $\dg(\hist)$ is the argument that the theory plays as a counter-argument to the path $\hist$.

Formally speaking, a defense relation $\dg \subseteq \allhist × \allargs$ is a right-unique relation satisfying bounded depth (to be defined shortly). Right-unique means that given any $\hist \in \allhist$, $\dg(\hist) = \set{\ar \in \allargs \suchthat (\hist, \ar) \in \dg}$ is empty or is a singleton.
As is usual for right-unique relations, we write $\dg(\hist) = \ar$ instead of $\dg(\hist) = \set{\ar}$ when $\dg(\hist)$ is a singleton (and $\dg(\hist) = \emptyset$ otherwise).
Given $H \subseteq \allhist$, define $\dg(H) = \set{\dg(\hist) \suchthat \alpha \in H} \subseteq \allargs$. 
Given $\ar_0, \ar_1 \in \allargs$, say that $\ar_1$ defends $\ar_0$ 
iff $\exists \hist \in \allhist \suchthat \ar_0 \in R(\hist) \land \dg(\hist) = \ar_1$, equivalently, 
iff $\ar_0 \in R(\dg^{-1}(\ar_1))$, equivalently, 
iff $\ar_1 \in \dg(\set{\hist \suchthat \ar_0 \in R(\hist)})$.

Say that $\gamma$ has depth at least $k \in \N^*$ iff $\exists (\alpha_i, 1 ≤ i ≤ k) \suchthat \forall 1 ≤ i ≤ k, d(\alpha_i) ≠ \emptyset \land \forall 1 ≤ i ≤ k - 1, \alpha_i \cup \set{d(\alpha_i)} \subseteq \alpha_{i + 1}$.
The bounded depth condition on the defense relation imposes that for some $k \in \N^*$, $\gamma$ does not have depth at least $k$.
The depth of a theory is defined as the smallest value $k \in \N$ such that it does not have depth at least $k + 1$.
For example, $\gamma$ has depth $0$ iff $\dg = \emptyset$; and $\gamma$ has depth $1$ iff $\exists \alpha_1 \suchthat d(\alpha_1) ≠ \emptyset \land \forall \alpha_1 \cup \set{d(\alpha_1)} \subseteq \alpha_2: d(\alpha_2) = \emptyset$.
